Discovering cell type motifs in connectomics data

### 2013-12-26 docs/paper.tex
updates to citations

\section*{Introduction}
Emerging techniques \autocite{Morgan2013,Zador2012} promise to
quantify the location of each neuron within a volume of interest,
along with its connections to all other neurons. Far exceeding the
capacity of neuroanatomists to trace small circuits, this leads to
enormous datasets that quantify aspects of the nervous system. The impact of
this kind of high-throughput transition has been seen before -- 
rise of sequencing techniques necessitated the
development of novel computational methods to understand genomic
structure, ushering in an era of bioinformatics as an independent
discipline \autocite{}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
has been prominent in many areas, ranging from sensory (e.g. retina,
\autocite{} via processing (e.g. neocortex) to movement (e.g. spinal
cord) \autocite{}. Indeed, such repeated patterns also exists in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to back
out underlying connectivity motifs.

@@

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include spectral methods
\autocite{},   community-based-detection   methods  \autocite{},   and
stochastic block  models \autocite{Nowicki2001}.  While these
approaches powerfully  incorporate the probabilistic nature of neural
connections \autocite{} they do not generally consider the known fact
that  the  probability  of  connections  between  neurons  depends  on
distance. Combining such probabilistic models with a spatial component
to model distance promises to  improve the detection of microcircuitry
from neural data.

@@

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: both the cell
types and their spatial patterns of interconnection. We apply it to
three computing systems: the mouse retina connectome \autocite{}, the
c. elegans connectome \autocite{}, and a ``connectome'' of a classical
microprocessor \autocite{James2010}. Note we do not explicitly
consider cellular morphology, absolute position (cell layer), or any
chemical or genetic information. Yet in all cases, we discover
cell-type to cell-type connectivity that depends on spatial distance
and these connectivity functions depend on the two involved cell
types. Comparing the cell types discovered by the algorithms with
those obtained by anatomists based on orthogonal information reveals a
high degree of agreement. We present a scalable approach to infer
microcircuitry from connectivity data.

@@

There have existed a range of previous approaches to the discovery
of neural microcircuitry \autocite{} These generally involve a great deal
of manual labor and ad-hoc determination of what constitutes a “type”
of cell -- to this day there are disagreements in the literature as to
the “true” types in the mammalian retina. Much as phylogenomics has
changed our understanding of animal ontologies, modern large scale
data will allow the efficient unbiased discovery of cell types and
circuits. The sheer amount of available data demands the introduction
of algorithmic approaches.

The development of automatic identification and quantification of cell
type may also provide a new computational phenotype for quantifying the effect
of disease, genetic interventions, and developmentally-experienced neural activity. Our method can in principle identify neuron-types across non-connected graphs, e.g. across animals. For example, the types of neurons in one animal can be associated with the types of neurons in another animal, in the same way as this is already possible through molecular markers \autocite{}. This could be particularly important if cell types appear that are due to properties of the stimuli and experience as opposed to the molecular properties of cells \autocite{}. This would allow comparative quantitative anatomy across animals, and the search for the causes of the ultimate connectivity.

@@

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
initialized from different random initial points in the state space,
and then pick the chain with the highest log likelihood. Early
experiments attempted to use the full ensemble of chains as an
estimate of the posterior but were abandoned due to the difficulty of
summarizing the posterior over such a complex state space, which is an
ongoing area of research in the probabilistic modeling community
\autocite{}.

@@

The adjusted rand index (ARI) is a measure of the similarity of two
different clusterings \autocite{Rand1971} -- two identical clusters
have an ARI of 1.0 while progressively more dissimilar clusters have
lower ARIs, becoming negative as the clustering gets anti-correlated.

>>>>

\section*{Introduction}
Emerging techniques \autocite{Morgan2013,Zador2012} promise to
quantify the location of each neuron within a volume of interest,
along with its connections to all other neurons. Far exceeding the
capacity of neuroanatomists to trace small circuits, this leads to
enormous datasets that quantify aspects of the nervous system. The impact of
this kind of high-throughput transition has been seen before -- 
rise of sequencing techniques necessitated the
development of novel computational methods to understand genomic
structure, ushering in an era of bioinformatics as an independent
discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
has been prominent in many areas, ranging from sensory (e.g. retina,
\autocite{Masland2001} via processing (e.g. neocortex \autocite{Mountcastle1997})
to movement (e.g. spinal cord) \autocite{Grillner2005}. Indeed, such
repeated patterns also exists in human-made computing circuits. It
remains an important challenge to develop algorithms to use anatomical
data, e.g. connectomics, to back out underlying connectivity motifs.

@@

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not generally consider the
known fact that the probability of connections between neurons depends
on distance. Combining such probabilistic models with a spatial
component to model distance promises to improve the detection of
microcircuitry from neural data.

@@

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: both the cell
types and their spatial patterns of interconnection. We apply it to
three computing systems: the mouse retina connectome \autocite{Helmstaedter2013}, the
c. elegans connectome \autocite{White1986}, and a ``connectome'' of a classical
microprocessor \autocite{James2010}. Note we do not explicitly
consider cellular morphology, absolute position (cell layer), or any
chemical or genetic information. Yet in all cases, we discover
cell-type to cell-type connectivity that depends on spatial distance
and these connectivity functions depend on the two involved cell
types. Comparing the cell types discovered by the algorithms with
those obtained by anatomists based on orthogonal information reveals a
high degree of agreement. We present a scalable approach to infer
microcircuitry from connectivity data.

@@

There have existed a range of previous approaches to the discovery
of neural microcircuitry \autocite{missing} These generally involve a great deal
of manual labor and ad-hoc determination of what constitutes a “type”
of cell -- to this day there are disagreements in the literature as to
the “true” types in the mammalian retina. Much as phylogenomics has
changed our understanding of animal ontologies, modern large scale
data will allow the efficient unbiased discovery of cell types and
circuits. The sheer amount of available data demands the introduction
of algorithmic approaches.

The development of automatic identification and quantification of cell
type may also provide a new computational phenotype for quantifying the effect
of disease, genetic interventions, and developmentally-experienced neural activity. Our method can in principle identify neuron-types across non-connected graphs, e.g. across animals. For example, the types of neurons in one animal can be associated with the types of neurons in another animal, in the same way as this is already possible through molecular markers \autocite{Brown2009}. This could be particularly important if cell types appear that are due to properties of the stimuli and experience as opposed to the molecular properties of cells \autocite{missing}. This would allow comparative quantitative anatomy across animals, and the search for the causes of the ultimate connectivity.

@@

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
initialized from different random initial points in the state space,
and then pick the chain with the highest log likelihood. Early
experiments attempted to use the full ensemble of chains as an
estimate of the posterior but were abandoned due to the difficulty of
summarizing the posterior over such a complex state space, which is an
ongoing area of research in the probabilistic modeling community
\autocite{missing}.

@@

The adjusted rand index (ARI) is a measure of the similarity of two
different clusterings \autocite{Hubert1985} -- two identical clusters
have an ARI of 1.0 while progressively more dissimilar clusters have
lower ARIs, becoming negative as the clustering gets anti-correlated.
